Supplemental Methods: No robust coexistence in a canonical model of plant-soil feedbacks

### 2021-10-27 paper/cover_letter/PSF_cover.tex
final edits and updated cover letter

\opening{Dear Editor,}

@@

Please find attached the manuscript, ``No robust coexistence in a canonical model of plant-soil feedbacks,'' by Zachary R.
Miller, Pablo Lech\'{o}n, and Stefano Allesina, which we would like to submit to \textit{Ecology Letters} as a Letter.

In recent decades, plant-soil feedbacks (PSFs) -- indirect interactions between plants mediated by changes to local soil microbiota and chemistry -- have emerged as a leading mechanism to explain widespread coexistence in natural plant communities. The study of PSFs, both in theory and in the field, has been guided by a modeling framework introduced by Bever, Westover, and Antonovics in a landmark 1997 paper. However, the now-canonical ``Bever model'' is formulated for only two species, while natural communities may contain tens or hundreds of co-occurring plants. Bridging this gap has been a crucial ongoing challenge for the field.

Here, we extend the Bever model to include an arbitrary number of species and use results from evolutionary game theory to thoroughly characterize coexistence in this extended model. Our central finding is that coexistence of more than two species is virtually impossible within this framework. We show that multi-species coexistence can only arise under highly artificial parameter combinations, and collapses when these conditions are violated even slightly. Such fine-tuned conditions never arise by chance, are not robust to inevitable biological variation, and cannot explain coexistence in real-world systems.

These surprising results have significant implications for plant community ecology. The Bever model undergirds the analysis and interpretation of PSF experiments across a wide array of systems, and our findings suggest that applying two-species predictions to richer communities will yield incorrect inference about plant coexistence. Abundant empirical evidence indicates that PSFs do play an important role in maintaining diversity in plant communities; the apparent contradiction between these observations and the model predictions calls for critical reevaluation of model assumptions and urgent exploration of alternative theoretical frameworks.

Our analysis illuminates a long-standing disconnect between the low diversity of a simple, but widely utilized, PSF model, and the high diversity of real-world ecosystems. By demonstrating that this well-known model is incompatible with multi-species coexistence mediated by PSFs, we believe our results will kickstart new interest and progress in the theory behind this rapidly advancing area of ecology. Publishing these findings in Ecology Letters will allow us to reach the broad range of theoreticians and experimentalists across community, plant, and soil ecology who work with and rely on PSF models.

>>>>

\opening{Dear Dr. Coulson,}

@@

Please find attached the manuscript, ``No robust coexistence in a canonical model of plant-soil feedbacks,'' by Zachary R.
Miller, Pablo Lech\'{o}n-Alonso, and Stefano Allesina, which we would like to submit to \textit{Ecology Letters} as a Letter.

In recent decades, plant-soil feedbacks (PSFs) -- indirect interactions between plants mediated by changes to local soil microbiota -- have emerged as a leading mechanism to explain widespread coexistence in plant communities. The study of PSFs, both in theory and in the field, has been guided by a modeling framework introduced by Bever, Westover, and Antonovics in a landmark 1997 paper. However, the now-canonical ``Bever model'' is formulated for only two species, while natural communities may contain tens or hundreds of co-occurring plants. Bridging this gap has been a crucial ongoing challenge for the field.

Here, we extend the Bever model to include an arbitrary number of species and use results from evolutionary game theory to thoroughly characterize coexistence in this extended model. Our central finding is that coexistence of more than two species is virtually impossible within this framework. We show that multi-species coexistence can only arise under highly artificial parameter combinations, and collapses when these conditions are violated even slightly. Such fine-tuned conditions never arise by chance, are not robust to inevitable biological variation, and cannot explain coexistence in real-world communities.

These surprising results have significant implications for plant community ecology. The Bever model undergirds the analysis and interpretation of PSF experiments across a wide array of systems, and our findings suggest that applying two-species predictions to richer communities will yield incorrect inference about plant coexistence. Abundant empirical evidence indicates that PSFs do play an important role in maintaining diversity in plant communities; the apparent contradiction between these observations and the lack of coexistence predicted by this model calls for critical reevaluation of model assumptions and urgent exploration of alternative theory.

\textbf{Statement of novelty.} Our analysis illuminates a long-standing disconnect between the low diversity of a simple, but widely utilized, PSF model, and the high diversity of real-world ecosystems. To our knowledge, this work is the first to characterize the dynamics of the Bever model in full generality, representing an advancement on recent efforts (e.g. Eppinga \textit{et al.}, \textit{Nature Ecology \& Evolution 2018}; Mack \textit{et al.}, \textit{PLoS ONE} 2019). By demonstrating that this well-known PSF model is incompatible with multi-species coexistence, we believe our results will kickstart new interest and progress in the theory behind this rapidly advancing area of ecology. Publishing our findings in \textit{Ecology Letters} will allow us to reach the broad range of theoreticians and experimentalists across community, plant, and soil ecology who work with and rely on PSF models.

We have not previously published research directly related to the Bever model for PSFs. In a recent manuscript (Miller \& Allesina 2021, \textit{in submission}), we introduce a mathematical model for a broad class of environmentally-mediated interactions, including PSFs as a typical example. This approach, which we cite in our present study, demonstrates how and when PSFs can facilitate the coexistence of many species, and represents an alternative framework for modeling these complex interactions.

### 2021-10-27 paper/paper.tex
included figure 3 in manuscript
final edits and updated cover letter

\noindent Author contributions: ZRM and SA performed the mathematical analysis, and PLA implemented the numerical simulations. ZRM wrote the manuscript, and all authors edited and discussed the text. \\

\noindent Data accessibility: Code for reproducing the numerical simulations is available at [ADD REPOSITORY].\\

\noindent Article type: Letter\\
Number of words in abstract: 137\\
Number of words in main text: \\
Number of references: 51 \\
Number of figures: 3

@@

\begin{abstract}
Plant-soil feedbacks (PSFs) are thought to represent a crucial mechanism generating frequency dependent dynamics in plant communities. Negative feedbacks, in particular, are routinely invoked to explain coexistence and the maintenance of diversity in species-rich communities. However, the primary modeling framework used to study PSFs considers only two plant species, and we lack clear theoretical expectations for how these complex interactions play out in communities with natural levels of diversity. Here, we demonstrate that this canonical model for PSFs is equivalent to a well-studied model from evolutionary game theory, and using this equivalence, we are able to characterize the dynamics with an arbitrary number of plant species. Surprisingly, we find that coexistence of more than two species is virtually impossible in this model, suggesting that alternative theoretical frameworks are needed to describe feedbacks observed in diverse natural communities. 
\end{abstract}

@@

A central finding of the analysis by Bever \textit{et al.} was that the denominator of $q_i^*$, which they termed the ``interaction coefficient'', $I_s$ ($= \alpha_{ii} - \alpha_{ij} + \alpha_{jj} - \alpha_{ji}$), controls the model dynamics: When $I_s > 0$, which represents a community with positive feedbacks, the equilibrium in Eq.~\ref{2spp_eq} is unstable, and the two species cannot coexist. On the other hand, when $I_s < 0$, the equilibrium is neutrally stable, and the dynamics cycle around it, providing a form of non-equilibrium coexistence. In fact, these conclusions depend on the existence of a feasible equilibrium (i.e. positive equilibrium values), which further requires that $\alpha_{ii} < \alpha_{ji}$, for both $i, j = 1, 2$, in order for the model to exhibit coexistence \citep{bever1997incorporating,ke2015incorporating}.

@@

Of course, most natural plant communities feature more than two coexisting species, and it is precisely in the most diverse communities where mechanisms of coexistence hold the greatest interest \citep{van2013plant}. While it is not immediately clear how to generalize Eq.~\ref{2spp_relative} to more than two species, Eq.~\ref{2spp_absolute} is naturally extended by maintaining the assumption that the overall growth rate for any plant is a weighted average of its growth rate in each soil type:

\begin{align} \label{nspp_absolute}
\begin{cases}
\frac{dx_i}{dt} &= x_i \left(\sum_{j} \alpha_{ij} \, q_j \right) \, , \quad  i = 1, \dots n \\
\frac{dy_i}{dt} &= y_i \left( \beta_i \, p_i \right) \, .
\end{cases}
\end{align}
From Eq.~\ref{nspp_absolute}, one can derive the $n$-species analogue of Eq.~\ref{2spp_relative} (Fig.~\ref{fig:concept}),

\begin{align} \label{nspp_relative}
\begin{cases}
\frac{dp_i}{dt} &= p_i \left(\sum_{j} \alpha_{ij} \, q_j - \sum_{j, k} \alpha_{jk} \, p_j \, q_k \right) \, , \quad  i = 1, \dots n \\
\frac{dq_i}{dt} &= q_i \left(\beta_{i} \, p_i - \sum_{j} \beta_{j} \, p_j \, q_j  \right) \,
\end{cases}
\end{align}
giving the dynamics for species and soil component frequencies. Eq.~\ref{nspp_relative} is conveniently expressed in matrix form as

@@

Another notable property of bimatrix game dynamics is that the vector field defined by the model equations is divergence-free or incompressible \citep[see][for a proof]{hofbauer1998evolutionary}. The divergence theorem from vector calculus \citep{arfken1985mathematical} then tells us that Eq.~\ref{matrix_form} cannot have any attractors -- that is, regions of the phase space that ``pull in'' trajectories -- with multiple species. This rules out coexistence in a stable limit cycle or other non-equilibrium attractors (e.g. chaotic attractors). Thus, only the relatively fragile coexistence afforded by neutral oscillations is possible, as in the two-species model.

@@

\begin{equation}
	H = (\alpha_{12} - \alpha_{22}) \log q_1 + (\alpha_{21} - \alpha_{11}) \log q_2 + \beta_2 \log p_1 + \beta_1 \log p_2 .
\end{equation} 
Using the chain rule and time derivatives in Eq.~\ref{2spp_relative}, it is easy to show that $\frac{dH}{dt} = 0$ for any plant and soil frequencies (see Supplemental Methods). The level curves of $H$ form closed orbits around the equilibrium when the equilibrium is neutrally stable. Thus, $H$ implicitly defines the trajectories of the model, and can be used to determine characteristics such as the amplitude of oscillations arising from particular initial frequencies.

Because neutral cycles provide the only possible form of coexistence in this model, a key question becomes whether and when neutral cycles with $n$ species can arise. Do the ``negative feedback'' conditions identified by Bever \textit{et al.} generalize in richer communities? Indeed, they do; however, for more than two species, these conditions are very severe. The model in Eq.~\ref{matrix_form} supports oscillations with $n$ species -- for any $n$ -- if matrices $A$ and $B$ satisfy a precise relationship (Fig.~\ref{fig:fragile}; see Supplemental Methods for details). In particular, the model parameters must satisfy the conditions $\alpha_{ij} = \gamma_i + \delta_j$ for some constants $\gamma_i, \delta_i$ in $i = 1, \dots, n$ (when $i \neq j$), and $\alpha_{ii} = \gamma_i + \delta_i- c \beta_i$ (where $c$ is a positive constant independent of $i$). In the language of bimatrix games, such systems are called \emph{rescaled zero-sum games} \citep{hofbauer1996evolutionary,hofbauer1998evolutionary}. It is a long-standing conjecture in evolutionary game theory that these parameterizations are the \emph{only} cases where $n$-species coexistence can occur \citep{hofbauer1996evolutionary,hofbauer2011deterministic}.

Ecologically, these conditions mean there is a fixed effect of each soil type and plant species identity, and the growth rate of plant $i$ in soil type $j$ is the additive combination of these two, with no interaction effects. The only exception is for plants growing in their own soil type, which must experience a fitness cost ($\gamma_i + \delta_i - \alpha_{ii}$) exactly proportional to the rate at which they promote growth of the soil type ($\beta_i$).  
These conditions clearly extend the intuitive notion that each plant must have a disadvantage in its corresponding soil type to allow coexistence. But the parameters of the model are constrained so strongly that we never expect to observe cycles with more than two species in practice. When $n > 2$, a great deal of fine-tuning is necessary to satisfy the conditions yielding a rescaled zero-sum game; the probability that random parameters will be suitable is infinitesimally small. We confirm this numerically with simulations shown in Fig.~\ref{fig:histograms}. Although $n$-species cycles are clearly possible (as in Fig.~\ref{fig:fragile}), for parameters drawn independently at random, communities always collapse to zero, one, or two species, regardless of the initial richness.

@@

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{figure_2.pdf}
	\caption[Final community sizes with varying initial richness]
	{Final community sizes with varying initial richness. We show the distribution of final richness (number of species, in red) for 500 \textbf{(???)} communities governed by the $n$-species Bever model, initialized with 2, 3, 5, or 6 plant species. Parameters $\alpha_{ij}$ and $\beta_{i}$  were sampled independently from a standard uniform distribution, $U[0,1]$. For each random parameterization at each level of initial richness, we integrated the dynamics of Eq.~\ref{matrix_form} until the system reached a periodic orbit or until only one species remained. In agreement with the conjecture that coexistence of more than two species is vanishingly unlikely, we found that, regardless of the number of species initially present, every community collapsed to a subset of one or two surviving species.}
	\label{fig:histograms}
\end{figure*}

\begin{figure*}
	\centering
	%\includegraphics[width=\textwidth]{figure_3.pdf}
	\caption[HERE GOES FIGURE 3]
	{LABEL FOR FIG 3}
	\label{fig:fragile}
\end{figure*}

@@

>>>>

\noindent Author contributions: ZRM and SA performed the mathematical analysis, and PLA implemented the numerical simulations. ZRM wrote the manuscript, and all authors edited and discussed the text.\\

\noindent Data accessibility: Code for reproducing the numerical simulations is available at\\ \url{https://github.com/pablolich/plant_soil_feedback}.\\

\noindent Article type: Letter\\
Number of words in abstract: 135\\
Number of words in main text: 3297\\
Number of references: 52 \\
Number of figures: 3

\pagebreak

@@

\begin{abstract}
Plant-soil feedbacks (PSFs) are thought to represent a crucial mechanism generating frequency dependent dynamics in plant communities. Negative feedbacks, in particular, are routinely invoked to explain coexistence and the maintenance of diversity in species-rich communities. However, the primary modeling framework used to study PSFs considers only two plant species, and we lack clear theoretical expectations for how these complex interactions play out in communities with natural levels of diversity. Here, we demonstrate that this canonical model for PSFs is equivalent to a well-studied model from evolutionary game theory, and we use this equivalence to characterize the dynamics with an arbitrary number of plant species. Surprisingly, we find that coexistence of more than two species is virtually impossible in this model, suggesting that alternative theoretical frameworks are needed to describe feedbacks observed in diverse natural communities. 
\end{abstract}

@@

A central finding of the analysis by Bever \textit{et al.} was that the denominator of $q_i^\star$, which they termed the ``interaction coefficient'', $I_s$ ($= \alpha_{ii} - \alpha_{ij} + \alpha_{jj} - \alpha_{ji}$), controls the model dynamics: When $I_s > 0$, which represents a community with positive feedbacks, the equilibrium in Eq.~\ref{2spp_eq} is unstable, and the two species cannot coexist. On the other hand, when $I_s < 0$, the equilibrium is neutrally stable, and the dynamics cycle around it, providing a form of non-equilibrium coexistence. In fact, these conclusions depend on the existence of a feasible equilibrium (i.e. positive equilibrium values), which further requires that $\alpha_{ii} < \alpha_{ji}$ for both $i, j = 1, 2$, in order for the model to exhibit coexistence \citep{bever1997incorporating,ke2015incorporating}.

@@

Of course, most natural plant communities feature more than two coexisting species, and it is precisely in the most diverse communities that mechanisms of coexistence hold the greatest interest \citep{van2013plant}. While it is not immediately clear how to generalize Eq.~\ref{2spp_relative} to more than two species, Eq.~\ref{2spp_absolute} is naturally extended by maintaining the assumption that the overall growth rate for any plant is a weighted average of its growth rate in each soil type:

\begin{align} \label{nspp_absolute}
\begin{cases}
\frac{dx_i}{dt} &= x_i \left(\sum_{j} \alpha_{ij} \, q_j \right) \, , \quad  i = 1, \dots n \\
\frac{dy_i}{dt} &= y_i \left( \beta_i \, p_i \right) \, .
\end{cases}
\end{align}
From Eq.~\ref{nspp_absolute}, one can derive the $n$-species analogue of Eq.~\ref{2spp_relative},

\begin{align} \label{nspp_relative}
\begin{cases}
\frac{dp_i}{dt} &= p_i \left(\sum_{j} \alpha_{ij} \, q_j - \sum_{j, k} \alpha_{jk} \, p_j \, q_k \right) \, , \quad  i = 1, \dots n \\
\frac{dq_i}{dt} &= q_i \left(\beta_{i} \, p_i - \sum_{j} \beta_{j} \, p_j \, q_j  \right) \,
\end{cases}
\end{align}
giving the dynamics for species and soil component frequencies (Fig.~\ref{fig:concept}). Eq.~\ref{nspp_relative} is conveniently expressed in matrix form as

@@

Another notable property of bimatrix game dynamics is that the vector field defined by the model equations is divergence-free or incompressible \citep[see][for a proof]{hofbauer1998evolutionary}. The divergence theorem from vector calculus \citep{arfken1985mathematical} then dictates that Eq.~\ref{matrix_form} cannot have any attractors -- that is, regions of the phase space that ``pull in'' trajectories -- with multiple species. This rules out coexistence in a stable limit cycle or other non-equilibrium attractors (e.g. chaotic attractors). Thus, only the relatively fragile coexistence afforded by neutral oscillations is possible, as in the two-species model.

@@

\begin{equation}
	H = (\alpha_{12} - \alpha_{22}) \log q_1 + (\alpha_{21} - \alpha_{11}) \log q_2 + \beta_2 \log p_1 + \beta_1 \log p_2 .
\end{equation} 
Using the chain rule and time derivatives in Eq.~\ref{2spp_relative}, it is easy to show that $\frac{dH}{dt} = 0$ for any plant and soil frequencies (see Supplemental Methods). The level curves of $H$ form closed orbits around the equilibrium when the equilibrium is neutrally stable. Thus, $H$ implicitly defines the trajectories of the model, and can be used to determine characteristics such as the amplitude of oscillations arising from particular initial frequencies \citep{volterra1926variazioni}.

Because neutral cycles provide the only possible form of coexistence in this model, a key question becomes whether and when neutral cycles with $n$ species can arise. Do the ``negative feedback'' conditions identified by Bever \textit{et al.} generalize in richer communities? Indeed, they do; however, for more than two species, these conditions are very severe. The model in Eq.~\ref{matrix_form} supports oscillations with $n$ species -- for any $n$ -- if matrices $A$ and $B$ satisfy a precise relationship (see Supplemental Methods for details). In particular, the model parameters must satisfy the conditions $\alpha_{ij} = \gamma_i + \delta_j$ for some constants $\gamma_i, \delta_i$ in $i = 1, \dots, n$ (when $i \neq j$), and $\alpha_{ii} = \gamma_i + \delta_i- c \beta_i$ (where $c$ is a positive constant independent of $i$). In the language of bimatrix games, such systems are called \emph{rescaled zero-sum games} \citep{hofbauer1996evolutionary,hofbauer1998evolutionary}. It is a long-standing conjecture in evolutionary game theory that these parameterizations are the \emph{only} cases where $n$-species coexistence can occur \citep{hofbauer1996evolutionary,hofbauer2011deterministic}.

Ecologically, these conditions mean there is a fixed effect of each soil type and plant species identity, and the growth rate of plant $i$ in soil type $j$ is the additive combination of these two, with no interaction effects. The only exception is for plants growing in their own soil type, which must experience a fitness cost ($\gamma_i + \delta_i - \alpha_{ii}$) exactly proportional to the rate at which they promote growth of their soil type ($\beta_i$). These conditions clearly extend the intuitive notion that each plant must have a disadvantage in its corresponding soil type to allow coexistence. But the parameters of the model are constrained so strongly that we never expect to observe cycles with more than two species in practice. When $n > 2$, a great deal of fine-tuning is necessary to satisfy the conditions yielding a rescaled zero-sum game; the probability that random parameters will be suitable is infinitesimally small. We confirm this numerically with simulations shown in Fig.~\ref{fig:histograms}. Although $n$-species cycles are clearly possible (as in Fig.~\ref{fig:fragile}), for parameters drawn independently at random, communities always collapse to one or two species, regardless of the initial richness.

@@

@@

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{figure_2.pdf}
	\caption[Final community sizes with varying initial richness]
	{Final community sizes with varying initial richness. We show the distribution of final richness (number of species, in red) for 5000 communities governed by the $n$-species Bever model, initialized with 2, 3, 5, or 6 plant species. Parameters $\alpha_{ij}$ and $\beta_{i}$ were sampled independently from a standard uniform distribution, $U(0,1)$. For each random parameterization at each level of initial richness, we integrated the dynamics of Eq.~\ref{matrix_form} until the system reached a periodic orbit or until only one species remained. In agreement with the conjecture that coexistence of more than two species is vanishingly unlikely, we found that regardless of the number of species initially present, every community collapsed to a subset of one or two surviving species.}
	\label{fig:histograms}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{figure_3.pdf}
	\caption[Coexistence of three or more species is not robust]{Coexistence of three or more species is not robust. It is possible to obtain neutrally stable oscillations with any number of species if the model parameters constitute a rescaled zero-sum game (see text for conditions). Here, for example, we show sustained oscillations with 4 species using fine-tuned parameters, $A_u$ (left). However, if we randomly perturb $A_u$ by a small amount to obtain new parameters, $A_p$, the dynamics quickly collapse to a two-species subset (right). Any slight perturbation is enough to disrupt coexistence; for this example, the parameters $A_u$ and $A_p$ are highly correlated (inset) and differ in value by less than 3\% on average.}
	\label{fig:fragile}
\end{figure*}

### 2021-10-27 paper/supp_methods.tex
edited my last name
final edits and updated cover letter

\date{}
\author{Zachary R. Miller$^{1*}$, Pablo Lech\'{o}n$^{1}$, and Stefano Allesina$^{1,2}$ \\
	\\
	\normalsize{$^{1}$Department of Ecology \& Evolution, University of Chicago, Chicago, IL, USA}\\
	\normalsize{$^{2}$Northwestern Institute on Complex Systems, Evanston, IL, USA}\\
	\\
	\normalsize{*corresponding author e-mail: zachmiller@uchicago.edu}\\
}

@@

\begin{align}
\begin{cases}
\frac{dx_i}{dt} &= x_i \left(\sum_{j} \alpha_{ij} \, q_j \right) \, , \quad  i = 1, \dots n \\
\frac{dy_i}{dt} &= y_i \left( \beta_i \, p_i \right)
\end{cases}
\end{align}
governing the time-evolution of plant abundances $x_i$ and soil components $y_i$, where $p_i = x_i / \sum_j x_j$, $q_i = y_i / \sum_j y_j$, and Greek letters denote nonnegative parameters. These equations capture the assumptions outlined by \citet{bever1997incorporating} for two species and extend them straightforwardly to any $n$ species. Following the approach of Bever \textit{et al.} for two species \citep[and consistent with other generalizations of this model,][]{kulmatiski2008plant,eppinga2018frequency}, we derive dynamics for the $p_i$ by applying the chain rule:

\begin{align} \label{derive_relative}
\begin{split}
	\frac{dp_i}{dt} &= \frac{d}{dt} \frac{x_i}{\sum x_j}\\  
	&= \frac{1}{\sum_j x_j} \frac{dx_i}{dt} - \frac{x_i}{(\sum_j x_j)^2} \sum_j \frac{dx_j}{dt} \\
	&= \frac{x_i}{\sum_j x_j} \left( \sum_{j} \alpha_{ij} \, q_j \right) - \frac{x_i}{\sum_j x_j} \left(\sum_j \frac{x_j}{\sum_k x_k} \sum_{l} \alpha_{jl} \, q_l \right)\\
	&= p_i \left(\sum_{j} \alpha_{ij} \, q_j -  \sum_{j, k} \alpha_{jk} p_j  q_k \right) \, .
\end{split}
\end{align}
This last expression is identical to the first line of Eq.~5 in the Main Text. The dynamics for $q_i$ can be derived in exactly the same way (using the definitions $\beta_{ii} = \beta_i$ and $\beta_{ij} = 0$). The two terms of each per capita growth rate in Eq.~5 have natural interpretations in the language and notation of linear algebra: $\sum_{j} \alpha_{ij} \, q_j$ is the $i$th component of the matrix-vector product $A \bm{q}$ and $\sum_{j, k} \alpha_{jk} p_j  q_k$ is the bilinear form $\bm{p}^T A \bm{q}$. Here, $A$ (and $B$) is an $n \times n$ matrix and $\bm{p}$ ($\bm{q}$) is a vector of length $n$, as described in the Main Text. We can re-write Eq.~5 as
\begin{align}
\begin{split}
	\frac{dp_i}{dt} &= p_i \left((A \bm{q})_i - \bm{p}^T A \bm{q} \right) \\
	\frac{dq_i}{dt} &= q_i \left((B \bm{p})_i - \bm{q}^T B \bm{p} \right) \\
\end{split}
\end{align}
or even more compactly as
\begin{align}
\begin{cases}
\frac{d\bm{p}}{dt} &= D(\bm{p}) \left(A \bm{q} - (\bm{p}^T A \bm{q}) \bm{1} \right) \\
\frac{d\bm{q}}{dt} &= D(\bm{q}) \left(B \bm{p} - (\bm{q}^T B \bm{p}) \bm{1}  \right) \,
\end{cases}
\end{align}
which is Eq.~6 in the Main Text.

@@

\begin{align} \label{firstblock}
\begin{split} 
	\frac{\partial}{\partial p_j} \frac{dp_i}{dt} &= p_i \left( - \sum_{k}^{n-1} \alpha_{jk} \, q_k + \sum_{k}^{n-1} \alpha_{nk} q_k - \alpha_{jn} g(\bm{q}) + \alpha_{nn} g(\bm{q}) \right) \\
	&= p_i \left( - \sum_{k}^{n} \alpha_{jk} \, q_k + \sum_{k}^{n} \alpha_{nk} q_k \right)\\
	&= 0
\end{split}
\end{align}
Here, we have used the fact that $A \bm{q^*} \propto \bm{1}$. Notice that, because the factors in parentheses in Eq.~\ref{reduced} are zero at equilibrium, these community matrix calculations are valid even for $i = j$.

@@

\begin{align}
	\begin{split}
	\frac{dH}{dt} &= \sum_i p_i^\star \frac{1}{p_i} \frac{dp_i}{dt} + c \sum_j q_j^\star \frac{1}{q_i} \frac{dq_i}{dt} \\
	&= \sum_i p_i^\star \left(\sum_{j} \alpha_{ij} \, q_j - \sum_{j, k} \alpha_{jk} \, p_j \, q_k \right) + c \sum_j q_j^\star \left( \beta_{i} \, p_i - \sum_{j} \beta_{j} \, p_j \, q_j \right) \\
	&= \sum_{i, j} \alpha_{ij} \, p_i^\star \, q_j - \sum_{j, k} \alpha_{jk} \, p_j \, q_k + c \sum_{i} \beta_{i} \, q_i^\star \, p_i - c \sum_{j} \beta_{j} \, p_j \, q_j \\
	&= \sum_{i, j} \alpha_{ij} \, (p_i^\star - p_i) \, q_j + c \sum_{i} \beta_{i} \, (q_i^\star - q_i) \, p_i 
	\intertext{Now, because $A = -c B^T$, we have}
	&= c \sum_i \beta_i \left( -(p_i^\star - p_i) \, q_i + (q_i^\star - q_i) \, p_i \right)\\
	&= c \sum_i \beta_i \left( -p_i^\star \, q_i + q_i^\star \, p_i \right)
	\intertext{and because $q_i^* = p_i^* = \frac{Z}{\beta_{i}}$, with $Z$ the normalizing constant,}
	&= c \, Z \, \sum_i (-q_i + p_i) \\
	&= 0
	\end{split}
\end{align}
In the last line, we use the fact that both sets of frequencies always sum to one.

Each orbit remains in the level set defined by the initial conditions,$(\bm{p}_0, \bm{q}_0)$:

@@

>>>>

\date{}
\author{Zachary R. Miller$^{1*}$, Pablo Lech\'{o}n-Alonso$^{1}$, and Stefano Allesina$^{1,2}$ \\
	\\
	\normalsize{$^{1}$Department of Ecology \& Evolution, University of Chicago, Chicago, IL, USA}\\
	\normalsize{$^{2}$Northwestern Institute on Complex Systems, Evanston, IL, USA}\\
	\\
	\normalsize{*corresponding author e-mail: zachmiller@uchicago.edu}\\
}

@@

\begin{align}
\begin{cases}
\frac{dx_i}{dt} &= x_i \left(\sum_{j} \alpha_{ij} \, q_j \right) \, , \quad  i = 1, \dots n \\
\frac{dy_i}{dt} &= y_i \left( \beta_i \, p_i \right)
\end{cases}
\end{align}
governing the time-evolution of plant abundances $x_i$ and soil components $y_i$, where $p_i = x_i / \sum_j x_j$, $q_i = y_i / \sum_j y_j$, and Greek letters denote nonnegative parameters. These equations capture the assumptions outlined by \citet{bever1997incorporating} for two species and extend them straightforwardly to any $n$ species. Following the approach of Bever \textit{et al.} for two species \citep[and consistent with other generalizations of this model, e.g.,][]{kulmatiski2008plant,eppinga2018frequency}, we derive dynamics for the $p_i$ by applying the chain rule:

\begin{align} \label{derive_relative}
\begin{split}
	\frac{dp_i}{dt} &= \frac{d}{dt} \frac{x_i}{\sum x_j}\\  
	&= \frac{1}{\sum_j x_j} \frac{dx_i}{dt} - \frac{x_i}{(\sum_j x_j)^2} \sum_j \frac{dx_j}{dt} \\
	&= \frac{x_i}{\sum_j x_j} \left( \sum_{j} \alpha_{ij} \, q_j \right) - \frac{x_i}{\sum_j x_j} \left(\sum_j \frac{x_j}{\sum_k x_k} \sum_{l} \alpha_{jl} \, q_l \right)\\
	&= p_i \left(\sum_{j} \alpha_{ij} \, q_j -  \sum_{j, k} \alpha_{jk} p_j  q_k \right) \, .
\end{split}
\end{align}
This last expression is identical to the first line of Eq.~5 in the Main Text. The dynamics for $q_i$ can be derived in exactly the same way (using the definitions $\beta_{ii} = \beta_i$ and $\beta_{ij} = 0$). The two terms of each per capita growth rate in Eq.~5 have natural interpretations in the language and notation of linear algebra: $\sum_{j} \alpha_{ij} \, q_j$ is the $i$th component of the matrix-vector product $A \bm{q}$ and $\sum_{j, k} \alpha_{jk} p_j  q_k$ is the bilinear form $\bm{p}^T A \bm{q}$. Here, $A$ (and $B$) is an $n \times n$ matrix and $\bm{p}$ ($\bm{q}$) is a vector of length $n$, as described in the Main Text. We can re-write Eq.~5 as
\begin{align}
\begin{split}
	\frac{dp_i}{dt} &= p_i \left((A \bm{q})_i - \bm{p}^T A \bm{q} \right) \\
	\frac{dq_i}{dt} &= q_i \left((B \bm{p})_i - \bm{q}^T B \bm{p} \right) \\
\end{split}
\end{align}
or even more compactly as
\begin{align}\label{compact_form}
\begin{cases}
\frac{d\bm{p}}{dt} &= D(\bm{p}) \left(A \bm{q} - (\bm{p}^T A \bm{q}) \bm{1} \right) \\
\frac{d\bm{q}}{dt} &= D(\bm{q}) \left(B \bm{p} - (\bm{q}^T B \bm{p}) \bm{1}  \right) \,
\end{cases}
\end{align}
which is Eq.~6 in the Main Text.

@@

\begin{align} \label{firstblock}
\begin{split} 
	\frac{\partial}{\partial p_j} \frac{dp_i}{dt} &= p_i \left( - \sum_{k}^{n-1} \alpha_{jk} \, q_k + \sum_{k}^{n-1} \alpha_{nk} q_k - \alpha_{jn} g(\bm{q}) + \alpha_{nn} g(\bm{q}) \right) \\
	&= p_i \left( - \sum_{k}^{n} \alpha_{jk} \, q_k + \sum_{k}^{n} \alpha_{nk} q_k \right)\\
	&= 0
\end{split}
\end{align}
Here, we have used the fact that $A \bm{q^\star} \propto \bm{1}$. Notice that, because the factors in parentheses in Eq.~\ref{reduced} are zero at equilibrium, these community matrix calculations are valid even for $i = j$.

@@

\begin{align}
	\begin{split}
	\frac{dH}{dt} &= \sum_i p_i^\star \frac{1}{p_i} \frac{dp_i}{dt} + c \sum_j q_j^\star \frac{1}{q_i} \frac{dq_i}{dt} \\
	&= \sum_i p_i^\star \left(\sum_{j} \alpha_{ij} \, q_j - \sum_{j, k} \alpha_{jk} \, p_j \, q_k \right) + c \sum_j q_j^\star \left( \beta_{i} \, p_i - \sum_{j} \beta_{j} \, p_j \, q_j \right) \\
	&= \sum_{i, j} \alpha_{ij} \, p_i^\star \, q_j - \sum_{j, k} \alpha_{jk} \, p_j \, q_k + c \sum_{i} \beta_{i} \, q_i^\star \, p_i - c \sum_{j} \beta_{j} \, p_j \, q_j \\
	&= \sum_{i, j} \alpha_{ij} \, (p_i^\star - p_i) \, q_j + c \sum_{i} \beta_{i} \, (q_i^\star - q_i) \, p_i 
	\intertext{Now, because $A = -c B^T$, we have}
	&= c \sum_i \beta_i \left( -(p_i^\star - p_i) \, q_i + (q_i^\star - q_i) \, p_i \right)\\
	&= c \sum_i \beta_i \left( -p_i^\star \, q_i + q_i^\star \, p_i \right)
	\intertext{and because $q_i^\star = p_i^\star = \frac{Z}{\beta_{i}}$, with $Z$ the normalizing constant,}
	&= c \, Z \, \sum_i (-q_i + p_i) \\
	&= 0
	\end{split}
\end{align}
In the last line, we use the fact that both sets of frequencies always sum to one.

Each orbit remains in the level set defined by the initial conditions, $(\bm{p}_0, \bm{q}_0)$:

@@

\section{Numerical Simulations}
To complement our analytical findings, we investigated the dynamics of many randomly parameterized communities using numerical simulations. In particular, we integrated  Eq.~\ref{compact_form} with 2, 3, 5, or 6 initial plant species and corresponding soil components. For each case, we sampled 5000 parameter sets at random and integrated the dynamics in Python using SciPy's (version 1.7.1) solve\_ivp function with the ``BDF'' method. We sampled non-singular payoff matrices $ A $ and $ B $ with each non-zero element drawn independently from the uniform distribution U(0, 1). For every choice of parameters, we integrated the system until a subset with $\leq 2$ species was reached (which occurred in all cases).
